Causal Bandits: Learning Good Interventions via Causal Inference

### 2016-05-19 causal-bandit/nips_2016/discussion.tex
minor cosmetic sentences and comment on assumption

Algorithm~\ref{alg:general} for general causal bandit problems adaptively estimates the reward for all allowable interventions $a \in \calA$ over $T$ rounds by sampling and applying interventions from a distribution $\eta$.
Theorem~\ref{thm:general} shows that this algorithm has (up to log factors) simple regret that is $\bigo{\sqrt{m(\eta)/T}}$ and that the parameter $m(\eta)$ is always less than $N$.
The value of $m(\eta)$ is a uniform bound on the variance of the reward estimators $\hat{\mu}_a$ and, intuitively, problems where all variables' values in the causal model ``occur naturally'' when interventions are sampled from $\eta$ will have low values of $m(\eta)$.

>>>>

Algorithm~\ref{alg:general} for general causal bandit problems adaptively estimates the reward for all allowable interventions $a \in \calA$ over $T$ rounds by sampling and applying interventions from a distribution $\eta$.
Theorem~\ref{thm:general} shows that this algorithm has (up to log factors) simple regret that is $\smash{\mathcal O(\sqrt{m(\eta)/T)}}$ and that the parameter $m(\eta)$ is always less than $N$.
The value of $m(\eta)$ is a uniform bound on the variance of the reward estimators $\hat{\mu}_a$ and, intuitively, problems where all variables' values in the causal model ``occur naturally'' when interventions are sampled from $\eta$ will have low values of $m(\eta)$.

### 2016-05-19 causal-bandit/nips_2016/experiments.tex
fixes figure fonts
puts experiemental results in single figure
adds missing y-axis label to experimental results
minor cosmetic sentences and comment on assumption

\begin{figure}
  \centering
  \begin{minipage}[b]{0.3\textwidth}
    \includegraphics[width=\textwidth]{exp_regret_vs_m_N50_T400_s10000_20160205_2230}
    \caption{Simple regret vs $m(\boldsymbol{q})$ for fixed horizon $T=400$ and number of variables $N = 50$}
    \label{fig:simple_vs_m}
  \end{minipage}
  \hfill
  \begin{minipage}[b]{0.3\textwidth}
    \includegraphics[width=\textwidth]{exp_regret_vs_T_N50_a4_s10000_20160204_1831}
    \caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and $\epsilon = \sqrt{\frac{N}{8T}}$}
    \label{fig:simple_vs_T_vary_epsilon}
  \end{minipage}
  \hfill
  \begin{minipage}[b]{0.3\textwidth}
    \includegraphics[width=\textwidth]{exp_regret_vs_T_N50_s10000_20160206_0533}
    \caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and fixed $\epsilon = .3$}
    \label{fig:simple_vs_T}
  \end{minipage}
\end{figure}

Figure \ref{fig:simple_vs_T_vary_epsilon} demonstrates the performance of the algorithms in the worst case environment for standard bandits, where the gap between the optimal and sub-optimal arms, $\epsilon = \sqrt{\frac{N}{8T}}$ , is just too small to be learned. This gap is learn-able by the causal algorithms, for which the worst case $\epsilon$ depends on $m << N$. In figure \ref{fig:simple_vs_T} we fix $N$ and $\epsilon$ and observe that, for sufficiently large $T$, the regret decays exponentially. The decay constant is larger for the causal algorithms as they have observed a greater effective number of samples for a given $T$.

>>>>

\begin{figure}
    \begin{subfigure}[t]{0.3\textwidth}
		\centering    
    		\includegraphics[width=\textwidth]{exp_regret_vs_m_N50_T400_s10000_20160519_1030.pdf}
    		\caption{Simple regret vs $m(\boldsymbol{q})$ for fixed horizon $T=400$ and number of variables $N = 50$}
        \label{fig:simple_vs_m}
    \end{subfigure}\hfill
    \begin{subfigure}[t]{0.3\textwidth}
    		\centering
        \includegraphics[width=\textwidth]{experiment2_N50_a4_s10000_20160519_1030.pdf}
    		\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and $\epsilon = \sqrt{\frac{N}{8T}}$}
        \label{fig:simple_vs_T_vary_epsilon}
    \end{subfigure}\hfill
    \begin{subfigure}[t]{0.3\textwidth}
    		\centering
    		\includegraphics[width=\textwidth]{experiment3_N50_s10000_20160519_1030.pdf}
    		\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and fixed $\epsilon = .3$}
    		\label{fig:simple_vs_T}
    \end{subfigure}
    \caption{Experimental results}
    \label{fig:experiments}
\end{figure}

%\begin{figure}
%  \centering
%  \begin{minipage}[b]{0.3\textwidth}
%    \includegraphics[width=\textwidth]{exp_regret_vs_m_N50_T400_s10000_20160518_1534.pdf}
%    \caption{Simple regret vs $m(\boldsymbol{q})$ for fixed horizon $T=400$ and number of variables $N = 50$}
%    \label{fig:simple_vs_m}
%  \end{minipage}
%  \hfill
%  \begin{minipage}[b]{0.3\textwidth}
%    \includegraphics[width=\textwidth]{experiment2_N50_a4_s10000_20160518_1534.pdf}
%    \caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and $\epsilon = \sqrt{\frac{N}{8T}}$}
%    \label{fig:simple_vs_T_vary_epsilon}
%  \end{minipage}
%  \hfill
%  \begin{minipage}[b]{0.3\textwidth}
%    \includegraphics[width=\textwidth]{experiment3_N50_s10000_20160518_1534.pdf}
%    \caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and fixed $\epsilon = .3$}
%    \label{fig:simple_vs_T}
%  \end{minipage}
%\end{figure}

Figure \ref{fig:simple_vs_T_vary_epsilon} demonstrates the performance of the algorithms in the worst case environment for standard bandits, where the gap between the optimal and sub-optimal arms, $\smash{\epsilon = \sqrt{N/(8T)}}$ , is just too small to be learned. This gap is learn-able by the causal algorithms, for which the worst case $\epsilon$ depends on $m \ll N$. In figure \ref{fig:simple_vs_T} we fix $N$ and $\epsilon$ and observe that, for sufficiently large $T$, the regret decays exponentially. The decay constant is larger for the causal algorithms as they have observed a greater effective number of samples for a given $T$.

### 2016-05-19 causal-bandit/nips_2016/simple-regret.tex
corrects definition of m(q)
minor cosmetic sentences and comment on assumption

The algorithm operates as follows. For the first $T/2$ rounds it chooses $do()$ to collect observational data. As the only link from each $X_1,\ldots,X_N$ to $Y$ is a direct, causal one, $\P{Y|do(X_i=j)}=\P{Y|X_i=j}$. Thus we can create good estimators for the returns of the actions $do(X_i = j)$ for which $\P{X_i = j}$ is large. The actions for which $\P{X_i = j}$ is small may not be observed (often) so  estimates their returns could be poor. To address this, the remaining $T/2$ rounds are evenly split to estimate the rewards for these infrequently observed actions. The difficulty of the problem depends on $\vec{q}$ and, in particular, how many of the variables are unbalanced (\ie, small $q_i$ or $(1-q_i)$). For $\tau \in [1...N]$ let $I_\tau = \set{ i : \min\set{q_i, 1-q_i} \leq \frac{1}{\tau - 1}}$. Define
\eq{
\label{eq:m-simple}
m(\vec{q}) = \min \set{ d : |I_{d}| \leq d}\,.
% m(\vec{q}) = \max \set{i : \sum_{k=1}^N \ind{\min\set{q_k, \,1 - q_k} \leq 1/i} \leq i}\,.
}
%Let $I_1 = \emptyset$ and for $\tau \in [0,1)$ let $I_\tau = \set{ i : \min\set{q_i, 1-q_i} \leq \tau}$.

@@

$I_\tau$ is the set of variables considered unbalanced and we tune $\tau$ to trade off identifying the low probability actions against not having too many of them, so as to minimize the worst-case simple regret. When $\vec{q} = (\frac{1}{2}, \ldots, \frac{1}{2})$ we have $m(\vec{q}) = 1$ and when $\vec{q} = (0, \ldots, 0)$ we have $m(\vec{q}) = N$. We do not assume that $\vec{q}$ is known, thus Algorithm \ref{alg:simple} also utilizes the samples captured during the observational phase to estimate $m(\vec{q})$. Although very simple, the following two theorems show that this algorithm is effectively optimal.

@@

\ifsup
The proofs of Theorems \ref{thm:uq-simple} and \ref{thm:lower} may be found in Sections \ref{sec:thm:uq-simple} and \ref{sec:thm:lower} respectively.
\else
The proofs of Theorems \ref{thm:uq-simple} and \ref{thm:lower} follow by carefully analysing the concentration
of $\hat p_a$ and $\hat m$ about their true values and may be found in the supplementary material.
\fi
%We prove a lower bound on the simple regret that matches up to logarithmic factors the upper bound given in Theorem \ref{thm:uq-simple}. 
By utilizing knowledge of the causal structure, Algorithm \ref{alg:simple} effectively only has to explore the $m(\vec{q})$ 'difficult' actions. Standard multi-armed bandit algorithms must explore all $2N$ actions and thus achieve regret  $\Omega(\sqrt{N/T})$. Since $m$ is typically much smaller than $N$, the new algorithm can significantly outperform classical bandit algorithms in this setting. In practice, you would combine the data from both phases to estimate rewards for the low probability actions. We do not do so here as it slightly complicates the proofs and does not improve the worst case regret.

>>>>

The algorithm operates as follows. For the first $T/2$ rounds it chooses $do()$ to collect observational data. As the only link from each $X_1,\ldots,X_N$ to $Y$ is a direct, causal one, $\P{Y|do(X_i=j)}=\P{Y|X_i=j}$. Thus we can create good estimators for the returns of the actions $do(X_i = j)$ for which $\P{X_i = j}$ is large. The actions for which $\P{X_i = j}$ is small may not be observed (often) so  estimates their returns could be poor. To address this, the remaining $T/2$ rounds are evenly split to estimate the rewards for these infrequently observed actions. The difficulty of the problem depends on $\vec{q}$ and, in particular, how many of the variables are unbalanced (\ie, small $q_i$ or $(1-q_i)$). For $\tau \in [2...N]$ let $I_\tau = \set{ i : \min\set{q_i, 1-q_i} < \frac{1}{\tau}}$. Define
\eq{
\label{eq:m-simple}
m(\vec{q}) = \min \set{ \tau : |I_{\tau}| \leq \tau}\,.
% m(\vec{q}) = \max \set{i : \sum_{k=1}^N \ind{\min\set{q_k, \,1 - q_k} \leq 1/i} \leq i}\,.
}
%Let $I_1 = \emptyset$ and for $\tau \in [0,1)$ let $I_\tau = \set{ i : \min\set{q_i, 1-q_i} \leq \tau}$.

@@

$I_\tau$ is the set of variables considered unbalanced and we tune $\tau$ to trade off identifying the low probability actions against not having too many of them, so as to minimize the worst-case simple regret. When $\vec{q} = (\frac{1}{2}, \ldots, \frac{1}{2})$ we have $m(\vec{q}) = 2$ and when $\vec{q} = (0, \ldots, 0)$ we have $m(\vec{q}) = N$. We do not assume that $\vec{q}$ is known, thus Algorithm \ref{alg:simple} also utilizes the samples captured during the observational phase to estimate $m(\vec{q})$. Although very simple, the following two theorems show that this algorithm is effectively optimal.

@@

\ifsup
The proofs of Theorems \ref{thm:uq-simple} and \ref{thm:lower} may be found in Sections \ref{sec:thm:uq-simple} and \ref{sec:thm:lower} respectively.
\else
The proofs of Theorems \ref{thm:uq-simple} and \ref{thm:lower} follow by carefully analysing the concentration
of $\hat p_a$ and $\hat m$ about their true values and may be found in the supplementary material.
\fi
%We prove a lower bound on the simple regret that matches up to logarithmic factors the upper bound given in Theorem \ref{thm:uq-simple}. 
By utilizing knowledge of the causal structure, Algorithm \ref{alg:simple} effectively only has to explore the $m(\vec{q})$ 'difficult' actions. Standard multi-armed bandit algorithms must explore all $2N$ actions and thus achieve regret  $\smash{\Omega(\sqrt{N/T})}$. Since $m$ is typically much smaller than $N$, the new algorithm can significantly outperform classical bandit algorithms in this setting. In practice, you would combine the data from both phases to estimate rewards for the low probability actions. We do not do so here as it slightly complicates the proofs and does not improve the worst case regret.
